---
title: "Cleaning Data With R"
author: "Paul G. Smith"
date: "10/31/2019"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cleaning Data With R
## Overview

The purpose of this project is to highlight issues related to cleaning data, and demonstrating many common strategies and techniques using R programming to clean and prep the data for analysis.

## Exploring raw data

Essential R commands:

- class, dim, names, str, glimpse, summary
- head, tail, print
- hist, plot
- gather -> combines multiple columns into two rows with key and value (tidyr)
- spread -> moves key value columns to multiple columns with keys as column names (tidyr)
- separate - > splits a column by _ or whatever seperator you choose to multiple columns (tidyr)
- unite -> combines multiple columns to one column with _ as the seperator (tidyr)
- as.character
- as.numeric
- as.integer
- as.factor
- as.logical
- ymd - (lubridate)
- mdy - (lubridate)
- hms - (lubridate)
- ymd_hms - (lubridate)

### What messy data looks like
- X column is just the row number
- days are spread but should be part of the date field and have one row per date
- year and month and day should be one data column
- measures are melted but should be each in a column
- and so on…
```{r echo=TRUE, message=TRUE, warning=TRUE}

# load and prep sample tables
weather <- readRDS('data/weather.rds')

# first look at a messy data set
# look at the first 6 rows
head(weather)
# look at the last 6 rows
tail(weather)  
# look at a summary of the data
str(weather)
```


### What clean data looks like
- Each row is a daily record of the weather
- There is a lot cleaned up here
- We will clean this data set in the last section

```{r echo=TRUE, message=TRUE, warning=TRUE}

# load and prep sample table
weather_clean <- readRDS('data/weather_clean.rds')

# look at the first 6 rows
head(weather_clean)
# look at the last 6 rows
tail(weather_clean)  
# look at a summary of the data
str(weather_clean)

```

### Exploring raw data

Using class, dim, names, str, glimpse, and summary

### Looking at the data

```{r echo=TRUE, message=TRUE, warning=TRUE}
# load the data
bmi <- readRDS('data/bmi.rds')

# look at the classgli
class(bmi)

# get the dimensions
dim(bmi)

# view the column names
names(bmi)
```

### Understanding the structure of the data

```{r echo=TRUE, message=TRUE, warning=TRUE}
# Check the structure of bmi
str(bmi)

# Load dplyr
library(dplyr)


# Check the structure of bmi, the dplyr way
glimpse(bmi)

# View a summary of bmi
summary (bmi)

```

### Look at the data

```{r echo=TRUE, message=TRUE, warning=TRUE}

# View the first 6 rows
head(bmi)

# View the first 15 rows
head(bmi, 15)

# View the last 6 rows
tail(bmi, 6)

# View the last 10 rows
tail(bmi, 10)
```

### Visualizing the data

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Histogram of BMIs from 2008
hist(bmi$Y2008)

# Scatter plot comparing BMIs from 1980 to those from 2008
plot(bmi$Y1980, bmi$Y2008)

```

## Tidying data

### Introduction to tidy data

- Hadley Wickham’s paper on this from 2014 is the best resource for understanding what is tidy data
- Essentially it mean one row per observation and one type of observation per table
  + Observations as rows
  + Variables as columns
  + One type of observational unit per table
- If you see values in the column names like months or years, or stock names, you have untidy data

### Introduction to tidyr

Could do a whole notebook on this package, it does a lot
Here we really just use:

- *gather* combines multiple columns into two rows with key and value
- *spread* moves key value columns to multiple columns with keys as column names
- *separate* splits a column by _ or whatever seperator you choose to multiple columns
- *unite* combines multiple columns to one column with _ as the seperator

### Gathering columns into key-value-pairs

```{r echo=TRUE, message=TRUE, warning=TRUE}

# load tidyr
library(tidyr)

# look at the first 6 rows
head(bmi)

# Apply gather() to bmi and save the result as bmi_long
bmi_long <- gather(bmi, year, bmi_val, -Country)

# View the first 20 rows of the result
head(bmi_long,20)

```

### Spreading key-value pairs

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=FALSE}

# Apply spread() to bmi_long
bmi_wide <- spread(bmi_long, year, bmi_val)

# View the head of bmi_wide
head(bmi_wide)
```

### Separating columns

```{r echo=TRUE, message=TRUE, warning=TRUE}
# load the bmi_cc file
bmi_cc <- readRDS('data/bmi_cc.rds')

# view the head
head(bmi_cc)

# use separate to split Country_ISO
bmi_cc_clean <- separate(bmi_cc, col = Country_ISO, 
                         into = c("Country", "ISO"), sep = "/")

# view the result
head(bmi_cc_clean)
```


### Uniting columns

```{r echo=TRUE, message=TRUE, warning=TRUE}
# use unite to join the two fields with a "-"
bmi_cc <- unite(bmi_cc_clean, Country_ISO, Country, ISO, sep = "-")

# View the results
head(bmi_cc)
```

### Addressing common symptoms of messy data

- Column headers are values, not variable names
  + eye colors as column names with boolean values in rows
- Variable are stored in both rows and columns
  + pet counts stored as key value pair in two columns. n_dogs and n_cats should be seperate columns
- Multiple variables are stored in one column
sex_age column with M.34, F.55 as values
A single observational unit is stored in multiple tables
Multiple types of observational units are stored in the same table
This should be split into two tables with key like most sql beginner data sets of sales and sales persons

### Column headers are values, not variable names


## Preparing data for analysis

### Type conversions

- common type conversions
  + as.character
  + as.numeric
  + as.integer
  + as.factor
  + as.logical
- also a log of date conversions in lubridate
  + ymd, mdy, hms, ymd_hms, etc
  
### Types of variables in R

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Make this evaluate to character
class(as.character('true'))

# Make this evaluate to numeric
class(as.numeric("8484.00"))

# Make this evaluate to integer
class(as.integer(99))

# Make this evaluate to factor
class(as.factor("factor"))

# Make this evaluate to logical
class(as.logical("FALSE"))

```

### Common type conversions

```{r echo=TRUE, message=TRUE, warning=TRUE}
# load student data
students <- readRDS('data/students.rds')

# look at the structure
str(students)

# Coerce Grades to character
students$Grades <- as.character(students$Grades)

# Coerce Medu to factor
students$Medu <- as.factor(students$Medu)

# Coerce Fedu to factor
students$Fedu <- as.factor(students$Medu) 
    
# Look at students once more with str()
str(students)

```

### Working with dates

```{r echo=TRUE, message=TRUE, warning=TRUE}
# load students2 data
students2 <- readRDS('data/students2.rds')

# Preview students2 with str()
str(students2)

# load lubridate
library(lubridate)

# Parse as date
dmy("17 Sep 2015")

# Parse as date and time (with no seconds!)
mdy_hm("July 15, 2012 12:56")

# Coerce dob to a date (with no time)
students2$dob <- ymd(students2$dob)

# Coerce nurse_visit to a date and time
students2$nurse_visit <- ymd_hms(students2$nurse_visit)
    
# Look at students2 once more with str()
str(students2)

```

### Trimming and padding strings

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Load stringr
library(stringr)

# Trim all leading and trailing whitespace
str_trim(c("   Filip ", "Nick  ", " Jonathan"))

# Pad these strings with leading zeros
str_pad(c("23485W", "8823453Q", "994Z"), width=9, side='left', pad="0")

```

### Upper and lower case

```{r echo=TRUE, message=TRUE, warning=TRUE}

states <- c("al", "ak", "az", "ar", "ca", "co", "ct", "de", "fl", "ga", 
"hi", "id", "il", "in", "ia", "ks", "ky", "la", "me", "md", "ma", 
"mi", "mn", "ms", "mo", "mt", "ne", "nv", "nh", "nj", "nm", "ny", 
"nc", "nd", "oh", "ok", "or", "pa", "ri", "sc", "sd", "tn", "tx", 
"ut", "vt", "va", "wa", "wv", "wi", "wy")

# Make states all uppercase and save result to states_upper
states_upper <- toupper(states)
states_upper

# Make states_upper all lowercase again
states_lower <- tolower(states_upper)
states_lower

```

### Finding and replacing strings

```{r}

# Look at the head of students2
head(students2)

# Detect all dates of birth (dob) in 1997
str_detect(students2$dob, "1997")

# In the sex column, replace "F" with "Female"...
students2$sex <- str_replace(students2$sex, "F", "Female")

# ...And "M" with "Male"
students2$sex <- str_replace(students2$sex, "M", "Male")

# View the head of students2
head(students2)

```

### Missing and special values

- There may be complete observations that are missing
  + This can happen if data ingestion process is interupted.
- NA, NaN, Inf, #N/A (Excel), single dot (SPSS,SAS), -1, -99999
- Could be anything really

### Finding missing values

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Load social data
social_df <- readRDS('data/social_df.rds')

# Use is.na() on the full social_df to spot all NAs
is.na(social_df)

# Use the any() function to ask whether there are any NAs in the data
any(is.na(social_df))

# Look at a summary() of the dataset
summary(social_df)

# use table() to look at the status column
table(social_df$status)

```

### Dealing with missing values

```{r}

# Replace empty strings in status with NA
social_df$status[social_df$status == ""] <- NA

# take a look at data again
social_df

# Use complete.cases() to see which rows have no missing values
complete.cases(social_df)

# Use na.omit() to remove all rows with any missing values
na.omit(social_df)

```

### Outliers and obvious errors

- Could be from data entry errors, or some measurement error
- *summary* and visualizations like *hist* or *boxplot* are good for finding these

### Dealing with outliers and obvious errors

```{r echo=TRUE, message=TRUE, warning=TRUE}

# load the student3 data
students3 <- readRDS('data/students3.rds')

# use summary to look at the data
summary(students3)

# use a histogram to help spot outliers in the age variable
hist(students3$age)

# Do the same thing with the absences variable
hist(students3$absences)

# Next make a histogram of absences, but force the zeros to be bucketed to the right of zero
hist(students3$absences, right=F)

```

### Another look at strange values

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Look at a boxplot of age
boxplot(students3$age)

# Look at a boxplot of absences
boxplot(students3$absences)

```

## Putting it all together

### Get a feel for the data
- First step is to look at and summmarize the data to see whats there

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Verify that weather is a data.frame
class(weather)

# Check the dimensions
dim(weather)

# Look at the column names
colnames(weather)

```

### Summarize the data

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Look at the structure of the data
str(weather)

# use glimpse() from dplyr to do the same
glimpse(weather)

# Summarize the data values
summary(weather)

```

### Take a closer look

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Check out the first 6 rows
head(weather)

# Look at the first 15 rows
head(weather, 15)

# Then look at the last 6 rows
tail(weather)

# tail() the last 10 rows
tail(weather, 10)

```

### Column names are values too
- Next steps are to tidy the data columns and rows

```{r echo=TRUE, message=TRUE, warning=TRUE}

## Load the tidyr package
library(tidyr)

# Gather the day columns
weather2 <- gather(weather, day, value, c(X1:X31), na.rm = TRUE)

# Look at the first six rows
head(weather2)

```

### Values are variable names

```{r echo=TRUE, message=TRUE, warning=TRUE}

# First remove column of row names
weather2 <- weather2[, -1]

# Then spread the measure data
weather3 <- spread(weather2, measure, value)

# look at the first 6 rowws
head(weather3)

```

### Clean up dates
- Next we clean up the variable types

```{r echo=TRUE, message=TRUE, warning=TRUE}

## The following packages are already loaded
## tidyr dplyr stringr lubridate

head(weather3)

# Remove X's from day column
weather3$day <- str_replace(weather3$day, "X","")

head(weather3[1:5])

# Unite the year, month, and day columns
weather4 <- unite(weather3, date, year, month, day, sep = "-")

head(weather4[1:5])

# Convert date column to proper date format using lubridates's ymd()
weather4$date <- ymd(weather4$date)

# look at the structure
str(weather4[1:5])

# Rearrange columns using dplyr's select()
weather5 <- select(weather4, date, Events, CloudCover:WindDirDegrees)

# Look at the first 6 rows
head(weather5)

```
### A closer look at column types

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Check out the structure of weather5
str(weather5)

# Examine the first 20 rows of weather5. Are most of the characters numeric?
head(weather5, 20)

# Try to convert PrecipitationIn to numeric
as.numeric(weather5$PrecipitationIn)

# Need to do this another way

```

### Column type conversions

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Replace T with 0 (T = trace)
weather5$PrecipitationIn <- str_replace(weather5$PrecipitationIn, "T", "0")

# Convert characters to numerics
weather6 <- mutate_each(weather5, funs(as.numeric), CloudCover:WindDirDegrees)

# Look at result
str(weather6)

```

### Find missing values
- Now we dig deeper to find missing values and errors

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Count missing values
sum(is.na(weather6))

# Find missing values
summary(weather6)

# Find indices of NAs in Max.Gust.SpeedMPH
ind <- which(is.na(weather6$Max.Gust.SpeedMPH))

# Look at the full rows for records missing Max.Gust.SpeedMPH
weather6[ind, ]

```

### An obvious error

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Look at the distributions for all of the variables
summary(weather6)

# Find row with Max.Humidity of 1000
ind <- which(weather6$Max.Humidity == 1000)

# Look at the data for that day
weather6[ind, ]

# Change 1000 to 100
weather6$Max.Humidity[ind] <- 100 

```

### Another obvious error

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Look at a summary of Mean.VisibilityMiles
summary(weather6$Mean.VisibilityMiles)

# Get index of row with -1 value
ind <- which(weather6$Mean.VisibilityMiles == -1) 

# Look at full row
weather6[ind,]

# Set Mean.VisibilityMiles to the appropriate value
weather6$Mean.VisibilityMiles[ind] <- 10

```

### Check other extreme values

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Run a summary of all of the data once more
summary(weather6)

# Make a histogram for MeanDew.PointF
hist(weather6$MeanDew.PointF)

# Now look at histogram for Min.TemperatureF
hist(weather6$Min.TemperatureF)

# Compare this to histogram for Mean.TemperatureF
hist(weather6$Mean.TemperatureF)

```

### Finishing touches

```{r echo=TRUE, message=TRUE, warning=TRUE}

# Create a vector of better column names
new_colnames <- c("date", "events", "cloud_cover", "max_dew_point_f", 
  "max_gust_speed_mph", "max_humidity", "max_sea_level_pressure_in", 
  "max_temperature_f", "max_visibility_miles", "max_wind_speed_mph", 
  "mean_humidity", "mean_sea_level_pressure_in", "mean_temperature_f", 
  "mean_visibility_miles", "mean_wind_speed_mph", "mean_dew_point_f", 
  "min_dew_point_f", "min_humidity", "min_sea_level_pressure_in", 
  "min_temperature_f", "min_visibility_miles", "precipitation_in", 
  "wind_dir_degrees"
)

# replace the names
names(weather6) <- new_colnames

# Replace empty cells in events column
weather6$events[weather6$events == ""] <- "None" 
    
# Look at the first 6 rows of weather6
head(weather6)

```

### Now your data are clean!

- each row is a single observation (day of year)
- each column is a variable
- date is a date type
- events are more clear and explicit
- missing values are known
- data errors have been found and fixed

