---
title: "Cleaning Data With R"
author: "Paul G. Smith"
date: "10/31/2019"
output: 
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cleaning Data With R
## Overview

The purpose of this project is to highlight issues related to cleaning data, and demonstrating many common strategies and techniques using R programming to clean and prep the data for analysis.

## Exploring raw data

Essential R commands:

- class, dim, names, str, glimpse, summary
- head, tail, print
- hist, plot
- gather -> combines multiple columns into two rows with key and value (tidyr)
- spread -> moves key value columns to multiple columns with keys as column names (tidyr)
- separate - > splits a column by _ or whatever seperator you choose to multiple columns (tidyr)
- unite -> combines multiple columns to one column with _ as the seperator (tidyr)
- as.character
- as.numeric
- as.integer
- as.factor
- as.logical
- ymd - (lubridate)
- mdy - (lubridate)
- hms - (lubridate)
- ymd_hms - (lubridate)

### What messy data looks like
- X column is just the row number
- days are spread but should be part of the date field and have one row per date
- year and month and day should be one data column
- measures are melted but should be each in a column
- and so on…
```{r echo=TRUE, message=TRUE, warning=TRUE}

# load and prep sample tables
weather <- read.csv("data/weather.csv", header=TRUE, sep=",", as.is=TRUE)

# first look at a messy data set
# look at the first 6 rows
head(weather)
# look at the last 6 rows
tail(weather)  
# look at a summary of the data
str(weather)
```


### What clean data looks like
- Each row is a daily record of the weather
- There is a lot cleaned up here
- We will clean this data set in the last section

```{r echo=TRUE, message=TRUE, warning=TRUE}

# load and prep sample table
weather_clean <- read.csv("./data/weather_clean.csv", header=TRUE, sep=",")
weather_clean[3:23] <- lapply(weather_clean[3:23], as.numeric)
weather_clean$events <- as.character(weather_clean$event)
weather_clean$date <- as.POSIXct(weather_clean$date, tz="", format("%m/%d/%Y"))

# look at the first 6 rows
head(weather_clean)
# look at the last 6 rows
tail(weather_clean)  
# look at a summary of the data
str(weather_clean)

```

### Exploring raw data

Using class, dim, names, str, glimpse, and summary

### Looking at the data

```{r echo=TRUE, message=TRUE, warning=TRUE}
# load the data
bmi <- read.csv("./data/bmi.csv", header=TRUE, sep=",")
bmi$Country <- as.character(bmi$Country)

# look at the class
class(bmi)

# get the dimensions
dim(bmi)

# view the column names
colnames(bmi)
```

### Understanding the structure of the data

```{r echo=TRUE, message=TRUE, warning=TRUE}
# look at the structure of the data set
str(bmi)

# load the dplyr library
library(dplyr)

# use dplyr to look at the structure
glimpse(bmi)

# view a summary of the variables
summary(bmi)

```

### Look at the data

```{r echo=TRUE, message=TRUE, warning=TRUE}

# look at the first 6 rows
head(bmi)

# look at the first 15 rows
head(bmi, 15)

# look at the last 6 rows
tail(bmi)

# look at the last 10 rows
tail(bmi, 10)
```

### Visualizing the data

```{r echo=TRUE, message=TRUE, warning=TRUE}

# histogram of BMIs from 2008
hist(bmi$Y2008)

# scatter plot comparing BMIs from 1980 to those from 2008
plot(bmi$Y1980, bmi$Y2008)
```

## Tidying data
## Preparing data for analysis

